Saint Helena Island Implementation

### 2025-03-11 Implementation doc/Draft1.Rmd
Removidos arquivos antigos e atualizadas imagens

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.align = "left")

@@

models <- StrathE2E2::e2e_ls("Files/")

model <- StrathE2E2::e2e_read("Saint_Helena_MA", "2010-2019-CNRM-ssp126", models.path = "Files")

Area <- read.csv(str_glue("Files/Saint_Helena_MA/2010-2019-CNRM-ssp126/Param/event_timing_SAINT_HELENA_MA_2010-2019-CNRM-ssp126.csv")) %>% 
  filter(str_detect(Description, "km2")) %>% 
  .$Value

@@

This document describes the configuration of StrathE2E for the Saint Helena Island off the South Mid-Atlantic Ridge and its parameterisation to enable stationary state fitting for both the baseline period (2010-2019) and future projections (2020-2069).
These represent contrasting periods of environmental conditions.

@@

```{r guilds, echo=FALSE, fig.cap="Ecological guilds of the StrathE2E model (Heath, 2021)", out.width = '100%'}
knitr::include_graphics("img/guilds.png")
```

@@

Santa Helena (15.96°S, 5.70°W) is a remote volcanic island, part of the British Overseas Territory of Saint Helena, Ascension, and Tristan da Cunha, with its top 800m rising above sea level (Carleton et al., 2010).
In 2016, Saint Helena was designated as a sustainable-use Marine Protected Area (MPA; Johnson et al., 2019) that encompasses its entire Exclusive Economic Zone (EEZ; Booth and Azar, 2009).

@@

[Habitat maps inshore for Saint Helena to be defined]

As of V.4, StrathE2E2 can represent an offshore "overhang" where open ocean does not contact the seafloor (Figure \@ref(fig:overhang)).
The perimeter of the offshore zone in the model domain is the edge of the MPA, with a "false bottom" (overhang) that exchanges with the deep sea.
By using the MPA to define the model domain, we accurately represent the protected area status of Saint Helena Island, which also corresponds to the Exclusive Economic Zone (EEZ).
The total area of the model is 450,906.82 km².

```{r domain, echo=FALSE, fig.cap="Map of the model domain. StrathE2E defines seabed sediment habitats as inshore (light gray) or offshore/overhang (dark gray). Within each zone, three sediment classes can be represented -- fine (muddy, 1), medium (sandy, 2) and coarse (gravel, 3). A fourth class (rock, 0) represents an absence of soft sediment. Note that the majority of the Saint Helena model domain is comprised by the 'overhang', which does not contact the sea floor.", out.width = '100%'}
knitr::include_graphics("img/domain.png")
```

@@

The depth boundary between deep and shallow layers was determined using vertical diffusivity values from NEMO-MEDUSA [@yool13] and mixed layer depth from Globcolour (www.globcolour.info/).
The shallow-deep layer division was set at 60 m, with a bottom depth of 600 m for the deep layer.
The offshore zone then extends to the edge of the EEZ, but with a "false bottom" that exchanges with the deep sea (without seafloor).
In the coastal zone, the polygon is defined for areas shallower than 60 m or within 4.5 km of the coast.

#### Parameters for relationship between median grain size, sediment porosity and permeability. Permeability is used as the basis for estimating hydraulic conductivity which is a parameter in the representation of sediment processes in the model: {.unnumbered}

@@

These relationships are coded into the StrathE2E2 R-package with the parameters in the csv setup file for the Saint Helena model.
The parameters are probably a reasonable starting point for any future model of a new region.
Derivation of the parameters is described in the following text sub-sections.

@@

tab_area_proportions_cap <- table_nums(name = "tab_area_proportions", 
    caption = stringr::str_glue("Area-proportions of the inshore and offshore zones and the thicknesses of the                                        water column layers. The sea surface area of the model domain is an estimated                                         {Area} km^2^."))
```

[Copied from Ascension, still needs compilation for Saint Helena]

@@

tibble(Property = c("Sea-surface area proportion",
                        "Upper layer thickness (m)",
                        "Lower layer thickness (m)"),
           `Inshore/shallow` = c(model[["data"]][["physical.parameters"]][["x_shallowprop"]], 
                                 model[["data"]][["physical.parameters"]][["si_depth"]], 
                                 NA),
           `Offshore/deep` = c(1 - model[["data"]][["physical.parameters"]][["x_shallowprop"]], 
                               model[["data"]][["physical.parameters"]][["so_depth"]], 
                               model[["data"]][["physical.parameters"]][["d_depth"]])) %>% 
  gt() %>% 
  fmt_number(columns = c("Offshore/deep", "Inshore/shallow"), decimals = 4, use_seps = FALSE)

@@

tab_habitat_proportions_cap <- table_nums(name = "tab_habitat_proportions", 
    caption = stringr::str_glue("Area proportions and other characteristics of the seabed habitat classes defined in the model by depth, rock or sediment type. The sea surface area of the model domain is an estimated {Area} km^2^. Grain size is the median in mm, Permeability in units of m^2^, nitrogen content in %dw."))
```

@@

physical <- read.csv(str_glue("Files/Saint_Helena_MA/2010-2019-CNRM-ssp126/Param/physical_parameters_ASCENSION.csv"))

@@

[Copied from SBS, still needs compilation for SPSP]

To calculate the natural disturbance rate, we needed depth-averaged current speeds (including tidal influence) and wave data.
We used data products from Copernicus Marine Data; however, only surface tide components were available instead of depth-averaged values, and the depth-averaged current speeds did not include tides.
Therefore, we approximated the influence of tides on the depth profile by assuming that the extra surface speed diminishes with depth at a certain decay rate.
We employed a rough approximation for non-stratification and a flat bottom from the National Oceanography Centre (NOC).
The equation assumes a constant speed in the upper half of the water column and attenuation in the lower half due to bottom friction, with speed decaying to zero at the bottom.

$$U_{} = U_{\text{current}} + \frac{U_{\text{tides }}}{1.07}$$ $$V_{} = V_{\text{current}} + \frac{V_{\text{tides }}}{1.07}$$

We used the Copernicus toolbox for subsetting the specific variables, the SBS domain, and the desired period via Command Line Interface (CLI), then post-processed the data in R.
Since the Copernicus Marine Data Store only hosts recent data for public download, we used data from 2022, the oldest full annual cycle available with all three variables.
Details about the products used below.

Waves: We used the Global Ocean Waves Analysis product (DOI: 10.48670/moi-00017) with a 1/12-degree resolution.
The parameters used every 3 hours were: significant wave height, mean wave direction, and primary swell wave period.
Specific variables included Sea surface wave significant height (VHM0, m), Sea surface wave direction (VMDR, °), and Sea surface primary swell wave mean period (VTM01SW1, s).

Currents: We used the Global Ocean Physics Analysis and Forecast product (DOI: 10.48670/moi-00016) with a 1/12-degree resolution.
The 6-hourly parameters used were Eastward sea water velocity (uo, m/s) and Northward sea water velocity (vo, m/s).

Tides: We used the Global Ocean Physics Analysis and Forecast product (DOI: 10.48670/moi-00016) with a 1/12-degree resolution.
The 1-hourly parameters used were Surface sea water x velocity due to tide (utide, m/s) and Surface sea water y velocity due to tide (vtide, m/s).

Even though the grid is shared, there were coastal cells with currents but missing data for waves and tides.
We interpolated missing coastal cells using the Voronoi tessellation method (nearest neighbor).
We also interpolated the depth-averaged currents and waves to match the hourly timestep of the tides using the `approx` function.

We combined all data into a single dataset and applied functions from the `bedshear` package to calculate bed shear stress for the six habitat types.

@@

$$log_{10}(porosity) = p_3 + p_4\left(\frac{1}{1+e^{(\frac{-log_{10}(D_{50})-p_1}{p_2})}}\right)$$

We use this relationship to calculate porosity for sea bed sediments in the Saint Helena (`r table.ref("tab_habitat_proportions")`), using an alternative parameterisation to Wilson [@pace21].
This alternative set of parameters extends the relationship to fine, muddy sediments (`r table.ref("tab_porosity")`).

@@

tab_porosity_cap <- table_nums(name = "tab_porosity", 
    caption = "The four parameters for the function relating sediment porosity to median grain size. From Pace et al. (in review)")
```

@@

The magnitude of the static (refactory) organic nitrogen detritus pool in each sediment type is a required input to the model.
The code includes an option to impute values from empirical relationships between total organic nitrogen (TON) and mud content, and between mud content and median grain size.
This relationship has been documented in the North Sea implementation [@heath21].
For the Saint Helena implementation, the total nitrogen content is internally calculated and based on that relationship.

@@

We gathered the spawning and recruitment periods for each guild based on literature regarding the main species, described below. In the model the annual weight-specific fecundity is assumed to be shed uniformly between the start and end dates of spawning. The annual cohort of larvae/juveniles of each fish and benthos guild is assumed to recruit to the settled stage at a uniform daily rate between the start and end dates.

@@

We found no specific data on the reproductive period of planktivorous species in Saint Helena, except for the report that chub mackerel (*Scomber japonicus*) spawns there in March (Edwards, 1990).
Therefore, we inferred a spawning period during the austral spring and summer, with recruitment occurring in autumn and winter, based on data from small pelagic species at the Saint Peter and Saint Paul Rocks (refer to the SPSP model implementation) and other regions of the South Atlantic.
The spawning period for the planktivorous guild was set from December 1st to March 31st, and recruitment from April 1st to June 30th.

@@

The reproductive period for the demersal guild was determined based on the spawning patterns of the rock hind grouper (*Epinephelus adscensionis*), a species of ecological and fisheries significance in Saint Helena Island.
Most species within the subfamily Epinephelinae form spawning aggregations and have extended spawning periods during the winter and spring months (Edwards, 1990).
On Saint Helena and Ascension Island, the spawning season spans from June to November, with a peak in August, coinciding with the austral winter (Edwards, 1990; Ascension Island Government, 2016; Nolan et al., 2017).
Fishers in Saint Helena also report that grouper are difficult to catch in winter months, another indicator of the spawning period (Riley et al., 2020).
In the absence of specific recruitment data, we set the recruitment period to start 90 days after the end of spawning and continue for 150 days.

@@

Reports from the Saint Helena Island government indicate that lobsters enter early reproductive stages from October 1st to December 31st, with a closed fishing season from January to the end of March due to the advanced reproductive stages of the brown spiny lobster (*Panulirus echinatus*) and stump lobster (*Scyllarides obtusus*) (St Helena, 2012; 2022).

@@

#### Deposit and filter benthos {.unnumbered}

Research on the reproductive activities of deposit and filter benthos in the Saint Helena region is limited.
We adopted the same spawning and recruitment periods used for the DF benthos guild in the South Brazil Bight implementation (See X source): spawning period from September 1st to February 1st and recruitment from August 1st to January 1st.

@@

```{r tab_migrant, fig.cap = tab_event_cap}

data_frame(Parameter = model[["data"]][["biological.events"]][["Description"]] %>% 
                        str_replace_all("_", " "),
           Value = model[["data"]][["biological.events"]][["Value"]]) %>% 
  .[1:16,] %>% 
  gt() %>%
  fmt_number(
    decimals = 0,
    use_seps = FALSE
  )

@@

```{r tuna, echo=FALSE, out.width='100%'}
knitr::include_graphics("img/tuna.png")
```

## Event timing parameters (fitted) {.unnumbered}

Saint Helena likely serves as a feeding ground for sub-adult fish, which migrate to spawning grounds upon reaching maturity (Wright et al., 2020).
Bigeye tuna (Thunnus obesus) and yellowfin tuna (Thunnus albacares) primarily spawn in tropical waters, such as the Gulf of Guinea and the Gulf of Mexico, during the summer (Carleton et al., 2010).
Recent tagging studies on yellowfin tuna show that they remain in Saint Helena waters for extended periods, ranging from six months to a year, during which they feed and grow (Wright et al., 2020).
They are typically recruited into the fishery between February and April each year and leave the population upon reaching a length of 120 cm (Wright et al., 2020).
Yellowfin tuna reach an average size of 96 to 120 cm at Cardno Seamount in May/June, suggesting migration out of the EEZ at this size (Wright et al., 2020; Wright et al., 2021).
Additionally, yellowfin tuna exhibit high fidelity to the area and remain in St. Helena waters for extended periods due to elevated productivity levels associated with the island and seamounts (White et al., 2007; Wright et al., 2021; Wright et al., 2019).

We utilized outputs from the SEAPODYM model, employed to characterize the movement dynamics of two tuna species in the Atlantic Ocean, as part of research conducted within the Mission Atlantic project [Merillet et al.
X (SOURCE)].
These outputs are based on the same four climate projections derived from NEMO-ERSEM physical outputs used in the StrathE2E driving data (see section X).
The projections are driven by two SSPs (SSP3-7.0 and SSP1-2.6) and two CMIP6 models (CNRM-CM6-1-HR and GFDL-ESM4).
Skipjack tuna (Katsuwonus pelamis) and albacore tuna (Thunnus alalunga) were modeled using SEAPODYM-MASS, an age-structured dynamic population model that accounts for prey availability.
Historical data were provided from 1990 to 2014, along with projections from 2015 to 2070, on a monthly basis and averaged across the Saint Helena model domain.
Data included biomass estimates for each SSP, model, and total stock biomass in the Atlantic.

@@

To identify the start and end days for annual immigration and emigration periods, we interpolated data to obtain daily values averaged by decade for each projection.
To synchronize with the StrathE2E annual cycle, each month was standardized to 30 days.
Seasonal cycles for each decade and projections were extracted using the decompose function in R, excluding trends and random components - for more information check the [function documentation](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/decompose/).
Immigration was defined as the period during which biomass increased from the 20th to the 80th percentile, while emigration was the period during which biomass decreased from the 80th to the 20th percentile.

@@

As the migratory guild in Saint Helena Island includes species beyond skipjack and albacore tunas, we utilized SEAPODYM outputs to parameterize the local biomass.
This was done using a scaling factor to account for the multi-species migratory biomass from the Atlantic, since local stock data are unavailable.
In addition to skipjack and albacore, we considered 16 other species contributing to migratory stocks in the Atlantic that may enter Saint Helena EEZ, including: wahoo (*Acanthocybium solandri*), frigate tuna (*Auxis thazard*), Galapagos shark (*Carcharhinus galapagensis*), common dolphinfish (*Coryphaena hippurus*), Atlantic sailfish (*Istiophorus albicans*), shortfin mako (*Isurus oxyrinchus*), white marlin (*Kajikia albida*), blue marlin (*Makaira nigricans*), blue shark (*Prionace glauca*), Atlantic bonito (*Sarda sarda*), smooth hammerhead (*Sphyrna zygaena*), yellowfin tuna (*Thunnus albacares*), southern bluefin tuna (*Thunnus maccoyii*), bigeye tuna (*Thunnus obesus*), Atlantic bluefin tuna (*Thunnus thynnus*), and swordfish (*Xiphias gladius*).

These migratory fish stocks in the Atlantic are managed under the International Convention for the Conservation of Atlantic Tunas (ICCAT), which relies on a consensus-driven decision-making process among member nations.
Biomass estimates for this study were derived from Sea Around Us (SAU) stock assessments, which utilize ICCAT data and the CMSY method (Pauly et al., 2020; see www.seaaroundus.org/cmsy-method for details).
These estimates covered the period from 2010 to 2018, and we used the averaged values for our calculations.
The combined biomass of the 18 key stocks in the Atlantic, which may enter the Ascension EEZ, totaled 5,503,844.45 tons.
This was treated as the ‘global’ stock biomass of migratory fish, with a portion assumed to enter the model domain in all decades and projections.

The scaling factor was derived by dividing this global biomass by the skipjack and albacore stocks provided by SEAPODYM.
This factor was then applied to local skipjack and albacore biomass from SEAPODYM to estimate the total migratory biomass for Saint Helena Island, averaged by decade, model, and SSP.
To quantify the proportion of Atlantic migratory biomass entering the model domain annually, we divided the scaled local biomass by the Atlantic migratory biomass.
To estimate the proportion of peak biomass that remained within the model domain without emigrating, we extracted the maximum and minimum values from the seasonal cycle.

Using these results, we derived the necessary parameters for lines 17–26 in the event_timing_SAINT_HELENA\_\*.csv files, for each decade and the four climate projections.

```{r tab_migrant_cap}

tab_migrantb_cap <- table_nums(name = "tab_migrantb", 
    caption = "Biological event timing parameters related to migratory fish constant across the time periods and climate projections")
```

`r table_nums("tab_migrantb")`

```{r tab_migrant_b, fig.cap = tab_migrantb_cap}

data_frame(Parameter = model[["data"]][["biological.events"]][["Description"]] %>% 
                        str_replace_all("_", " "),
           Value = model[["data"]][["biological.events"]][["Value"]]) %>% 
  .[17:20,] %>% 
  gt() %>%
  fmt_number(
    decimals = 3,
    use_seps = FALSE
  )
```

```{r migratory_baseline_cap}

migratory_baseline_cap <- table_nums(name = "migratory_baseline_cap", 
    caption = "Biological event timing parameters related to migratory fish for the baseline period (2010-2019) across the four climate projections.")
```

`r table_nums("migratory_baseline_cap")`

```{r migratory_baseline, fig.cap = migratory_baseline_cap}
migratory_fish_data <- data.frame(
  Parameter = c(
    "Propn of ocean popn entering domain each year",
    "Immigration start day",
    "Immigration end day",
    "Propn of peak popn in model domain which remains and does not emigrate",
    "Emigration start day",
    "Emigration end day"
  ),
  `SSP1-CNRM` = c("0.008001", "4", "145", "0.9590", "216", "291"),
  `SSP3-CNRM` = c("0.007891", "4", "110", "0.9608", "214", "291"),
  `SSP1-GFDL` = c("0.007757", "351", "107", "0.9466", "178", "280"),
  `SSP3-GFDL` = c("0.007892", "351", "109", "0.9553", "180", "280"),
  check.names = FALSE
)

migratory_fish_data %>%
  gt() %>%
  cols_align(align = "left", columns = vars(Parameter)) %>%
  cols_align(align = "center", columns = vars(`SSP1-CNRM`, `SSP3-CNRM`, `SSP1-GFDL`, `SSP3-GFDL`)) %>%
  cols_width(Parameter ~ px(220)) %>%
  tab_options(table.font.size = 15, table.align = "center")
```

>>>>

Sys.setlocale("LC_ALL", "pt_BR.UTF-8")
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.align = "left")

@@

model <- StrathE2E2::e2e_read("Saint_Helena_MA_MPA", "2010-2019-CNRM-ssp126", models.path = "Files")

Area <- read.csv(str_glue("Files/Saint_Helena_MA_MPA/2010-2019-CNRM-ssp126/Param/event_timing_SAINT_HELENA_MA_2010-2019-CNRM-ssp126.csv")) %>% 
  filter(str_detect(Description, "km²")) %>% 
  .$Value

@@

This document describes the configuration of StrathE2E for Saint Helena Island and its parameterisation to enable stationary state fitting for both the baseline period (2010-2019) and future projections (2020-2069). These represent contrasting periods of environmental conditions.

@@

```{r guilds, echo=FALSE, fig.cap="Ecological guilds of the StrathE2E model (Heath, 2021).", out.width = '100%'}
knitr::include_graphics("img/guilds.png")
```

@@

Saint Helena (15.96°S, 5.70°W) is a remote volcanic island with its top rising approximately 800 meters above sea level [@carleton10]. It is located off the South Mid-Atlantic Ridge and forms part of the British Overseas Territory of Saint Helena, Ascension, and Tristan da Cunha. The island has a 200-nautical-mile Exclusive Economic Zone (EEZ), which is governed and utilized by the St Helena Government and the local population. In 2016, Saint Helena's entire EEZ was designated as a IUCN Category VI sustainable-use Marine Protected Area (MPA) by the Environmental Protection Ordinance [@stg16].

Saint Helena does not have reef-building corals. Instead, its narrow inshore habitats consist of boulder and bedrock reefs, as well as sandy beaches [@stg23]. The majority of the EEZ comprises pelagic habitats and deep ocean waters.

@@

As of V.4, StrathE2E2 can represent an offshore "overhang" where open ocean does not contact the seafloor (Figure \@ref(fig:overhang)).
The perimeter of the offshore zone in the model domain is the edge of the MPA, with a "false bottom" (overhang) that exchanges with the deep sea.
By using the MPA to define the model domain, we accurately represent the protected area status of Saint Helena Island, which also corresponds to its EEZ.
The sea surface area of the model domain was estimated to be 450,906.82 km².

```{r domain, echo=FALSE, fig.cap="Map of the model domain. StrathE2E defines seabed sediment habitats as inshore (light gray) or offshore/overhang (dark gray). Within each zone, three sediment classes can be represented -- fine (muddy, 1), medium (sandy, 2) and coarse (gravel, 3). A fourth class (rock, 0) represents an absence of soft sediment. In this model implementation the inshore and offshore were divided into 25% sand, 25% rock and 50% gravel sediment, where most of the model domain (>99.7%) is overhang with a false seabed. The sea surface area of the model was estimated to be 450,906.82 km².", out.width = '100%'}
knitr::include_graphics("img/domain.png")
```

@@

The depth boundary between deep and shallow layers was determined using vertical diffusivity values from NEMO-MEDUSA [@yool13] and mixed layer depth from Globcolour [@globcolour].
The shallow-deep layer division was set at 60 m, with a bottom depth of 600 m for the deep layer. The offshore zone then extends to the edge of the EEZ, but with a “false bottom” that exchanges with the deep sea (without seafloor). 
In the coastal zone, the polygon is defined for areas shallower than 60 m or within 4.5 km of the coast. 
The sediment classes were roughly designated as sand, gravel, and rock based on the seabed habitats from the habitat map denoting seabed type in the Marine Management Plan [@stg23].

```{r sediment, echo=FALSE, out.width = '100%'}
knitr::include_graphics("img/island2.png")
```

\newpage
#### Parameters for relationship between median grain size, sediment porosity and permeability. Permeability is used as the basis for estimating hydraulic conductivity which is a parameter in the representation of sediment processes in the model: {.unnumbered}

@@

These relationships are coded into the StrathE2E2 R-package with the parameters in the csv setup file for the North Sea model.
The parameters are probably a reasonable starting point for any future model of a new region.
Derivation of the parameters is described in the following text sub-sections.

@@

tab_area_proportions_cap <- table_nums(name = "tab_area_proportions", 
    caption = stringr::str_glue("Area-proportions of the inshore and offshore zones and the thicknesses of the                                        water column layers. The sea surface area of the model domain is an estimated 450,906.82 km²."))
```

@@

data_frame(Property = c("Sea-surface area proportion",
                        "Upper layer thickness (m)",
                        "Lower layer thickness (m)"),
           `Inshore/shallow` = c(model[["data"]][["physical.parameters"]][["x_shallowprop"]], 
                                 model[["data"]][["physical.parameters"]][["si_depth"]], 
                                 NA),
           `Offshore/deep` = c(1 - model[["data"]][["physical.parameters"]][["x_shallowprop"]], 
                               model[["data"]][["physical.parameters"]][["so_depth"]], 
                               model[["data"]][["physical.parameters"]][["d_depth"]])) %>% 
  gt() %>% 
  fmt_number(columns = c("Offshore/deep", "Inshore/shallow"), decimals = 4, use_seps = FALSE)

@@

tab_habitat_proportions_cap <- table_nums(name = "tab_habitat_proportions", 
    caption = stringr::str_glue("Area proportions and other characteristics of the seabed habitat classes defined in the model by depth, rock or sediment type. The sea surface area of the model domain is an estimated 450,906.82 km². Grain size is the median in mm, Permeability in units of m^2^, nitrogen content in %dw."))
```

@@

physical <- read.csv(str_glue("Files/Saint_Helena_MA_MPA/2010-2019-CNRM-ssp126/Param/physical_parameters_SAINT_HELENA_MA.csv"))

@@

[based on Dougie notes from Silvia, need some review]

To estimate the natural disturbance rate, we applied a rudimentary linear model, as depth-averaged current speeds (including tidal influence) and wave data were unavailable at the local level. We compiled proportional disturbance rates from other regional StrathE2E implementations for mud, sand, and gravel sediments in inshore and offshore regions. Since disturbance is a proportion expected to decrease with depth, we plotted log-transformed disturbance against depth. While no clear pattern emerged, we identified an extreme outlier with unusually low disturbance at a shallow depth. As unaccounted factors likely influenced this outlier, we excluded it from the dataset. We then attempted to fit a linear model of the form log(disturbance) ∼ depth + inshore/offshore + sediment type. However, only the intercept and inshore/offshore variable were statistically significant. To refine the model, we applied stepwise AIC-based selection. This process suggested excluding sediment type, resulting in the final model log(disturbance) ∼ depth + inshore/offshore. All terms in the final model were statistically significant. Transforming the results back to disturbance (rather than log disturbance) and plotting them allowed us to approximate the disturbance vs. depth relationship for both inshore and offshore zones.

@@

$$log_{10}(porosity) = p_3 + p_4\left(\frac{1}{1+e^{(\frac{-log_{10}(D_{50})-p_1}{p_2})}}\right)$$
We use this relationship to calculate porosity for sea bed sediments in the Saint Helena (`r table.ref("tab_habitat_proportions")`), using an alternative parameterisation to Wilson [@pace21].
This alternative set of parameters extends the relationship to fine, muddy sediments (`r table.ref("tab_porosity")`).

@@

tab_porosity_cap <- table_nums(name = "tab_porosity", 
    caption = "The four parameters for the function relating sediment porosity to median grain size. From Pace et al. (2021)")
```

@@

The magnitude of the static (refractory) organic nitrogen detritus pool in each sediment type is a required input to the model.
The code includes an option to impute values from empirical relationships between total organic nitrogen (TON) and mud content, and between mud content and median grain size.
This relationship has been documented in the North Sea implementation [@heath21].
For the Saint Helena implementation, the total nitrogen content is internally calculated and based on that relationship.

@@

We gathered the spawning and recruitment periods for each guild based on literature regarding the main species, described below and in (`r table.ref("tab_event")`). In the model the annual weight-specific fecundity is assumed to be shed uniformly between the start and end dates of spawning. The annual cohort of larvae/juveniles of each fish and benthos guild is assumed to recruit to the settled stage at a uniform daily rate between the start and end dates.

```{r butterfly, echo=FALSE, out.width = '100%'}
knitr::include_graphics("img/butterfly.png")
```

@@

We found no specific data on the reproductive period of planktivorous species on Saint Helena, except for the report that chub mackerel (*Scomber japonicus*) spawns there in March [@edwards90].

Therefore, we inferred a spawning period during the austral spring and summer and recruitment in autumn and winter, also based on data from small pelagics at the Saint Peter and Saint Paul Rocks (SPSP, see more details in <https://www.marineresourcemodelling.maths.strath.ac.uk/strathe2e/articles/Implementations.html>). This was primarily derived from flyingfish (*Cypselurus cyanopterus*). In SPSP, spawning occurs from December to March [@lessa99; @pinheiro11], with recruitment occurring from April to June [@lessa04].

Based on this, the planktivorous guild’s spawning period was set from December 1 to March 31st, and recruitment from April 1st to June 30th.

@@

The reproductive period for the demersal guild was determined based on the spawning patterns of the rock hind grouper (*Epinephelus adscensionis*), a species of ecological and fisheries importance in Saint Helena Island.

Most species within the grouper subfamily Epinephelinae form spawning aggregations and spawn over extended periods during the winter and spring [@edwards90]. On Saint Helena and its closest neighbor, Ascension Island, the spawning season occurs from June to November, peaking in August, coinciding with the austral winter [@edwards90; @AIG16; @nolan17; @riley20].

Fishers in Saint Helena also report that grouper are difficult to catch in winter months, another indicator of the spawning period [@riley20]. In the absence of specific recruitment data, we set the recruitment period to start 90 days after the end of spawning and continue for 150 days.

@@

Reports from the Saint Helena Government indicate that lobsters enter early reproductive stages from October 1st to December 31st, with a closed fishing season from January to the end of March due to the advanced reproductive stages of the brown spiny lobster (*Panulirus echinatus*) and stump lobster (*Scyllarides obtusus*) [@sthelena22].

@@

#### Filter and deposit benthos {.unnumbered}

Research on the reproductive activities of filter and deposit feeding benthos in the Saint Helena is limited.
We adopted the same spawning and recruitment periods used for the FD benthos guild in the South Brazil Bight implementation (see more details at <https://www.marineresourcemodelling.maths.strath.ac.uk/strathe2e/articles/Implementation_The_Brazilian_Shelf.html>) - spawning period from September 1st to February 1st and recruitment from August 1st to January 1st.

@@

```{r tab_event_timing, fig.cap = tab_event_cap}

data_frame(Parameter = model[["data"]][["biological.events"]][["Description"]] %>% 
                        str_replace_all("_", " "),
           Value = model[["data"]][["biological.events"]][["Value"]]) %>% 
  .[1:16,] %>% 
  gt() %>%
  fmt_number(
    decimals = 0,
    use_seps = FALSE)

@@

```{r tuna, echo=FALSE, out.width='100%'}
knitr::include_graphics("img/tunas.png")
```

Saint Helena likely serves as a feeding ground for sub-adult fish, which migrate to spawning grounds upon reaching maturity [@wright20].
Bigeye tuna (*Thunnus obesus*) and yellowfin tuna (*Thunnus albacares*) primarily spawn in tropical waters, such as the Gulf of Guinea and the Gulf of Mexico, during the summer [@carleton10; @fao11].
Recent tagging studies on yellowfin tuna show that they remain in Saint Helena waters for extended periods, ranging from six months to a year, during which they feed and grow [@wright20].
They are typically recruited into the fishery between February and April each year and leave the population upon reaching a length of 120 cm [@wright20].
Yellowfin tuna reach an average size of 96 to 120 cm at Cardno Seamount in May/June, suggesting migration out of the EEZ at this size [@wright20; @wright21].

We utilized outputs from the SEAPODYM model to characterize the movement dynamics of tunas in the Atlantic Ocean, as part of research conducted within the Mission Atlantic project [@merillet24]. These outputs were based on the same four climate projections derived from NEMO-ERSEM physical outputs used in the StrathE2E driving data (see pages 26-29). The projections are driven by two SSPs (SSP3-7.0 and SSP1-2.6) and two CMIP6 models (CNRM-CM6-1-HR and GFDL-ESM4). Two species, skipjack tuna (*Katsuwonus pelamis*) and albacore tuna (*Thunnus alalunga*), were modeled using SEAPODYM-MASS, an age-structured dynamic population model that accounts for prey availability [@seapodym1; @seapodym2]. Historical data were provided from 1990 to 2014, along with projections from 2015 to 2070, on a monthly basis and averaged across the Saint Helena model domain. Data included local biomass estimates for each SSP variant, forcing model variant, and total stock biomass in the Atlantic.

@@

To identify the start and end days for annual immigration and emigration periods, we interpolated data to obtain daily values averaged by decade for each projection. To synchronize with the StrathE2E annual cycle, each month was standardized to 30 days. 
Seasonal cycles for each decade and projections were extracted using the decompose function in R, excluding trends and random components - for more information check the [function documentation](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/decompose/). 
Immigration and emigration periods were defined as the days when biomass increased from the 20th to 80th percentile and decreased from the 80th to 20th percentile, respectively.

@@

Discussions with the model’s authors suggested that parameterization favored skipjack biomass in tropical and equatorial regions. To address this, we scaled the local tuna biomass using a factor derived from regional stock assessment. The scaling equation below was applied in all Mission Atlantic case studies to adjusted SEAPODYM outputs to better reflect local stock assessments. Where no local assessments were available, as in the Saint Helena domain, α defaulted to 1. Additionally, another scaling factor (\(\lambda\)) was applied in cases where skipjack and albacore alone do not represent the entire migratory guild, as in the Saint Helena Island domain.

The scaled biomass was calculated as:

\[\alpha = \frac{\ln\left(\frac{x(t)}{X(t)}\right)}{\ln(P)}, \quad \lambda = \frac{Y(t)}{X(t)}\]

\[scaled \ biomass(t) = [P]^\alpha \times X(t) \times \lambda\]

Where:

-\(P\) is the proportion of the Atlantic stock entering the model domain annually, from SEAPODYM;\
-\(\alpha\) is the scaling factor to adjust the local biomass relative to known stock assessments;\
-\(X(t)\) is the Atlantic stock from SEAPODYM;\
-\(Y(t)\) is the Atlantic biomass from stock assessments;  
-\(x(t)\) is the local biomass from stock assessments;  
-\(\lambda\) is the scaling factor to the others Atlantic stocks that may enter the model domain

In the case of Saint Helena, \(\lambda\) was derived from stock assessments from the 2010s. Part of the management of Atlantic migratory fish stocks falls under the International Commission for the Conservation of Atlantic Tunas (ICCAT), which relies on a consensus-based decision-making process among its member nations. To estimate the ‘global’ stock biomass of migratory fish assumed to enter the model domain, we compiled data for 18 species, including key species such as the aforementioned skipjack, yellowfin, bigeye, and albacore tuna, along with wahoo (*Acanthocybium solandri*), frigate tuna (*Auxis thazard*), Galapagos shark (*Carcharhinus galapagensis*), common dolphinfish (*Coryphaena hippurus*), Atlantic sailfish (*Istiophorus albicans*), shortfin mako (*Isurus oxyrinchus*), white marlin (*Kajikia albida*), blue marlin (*Makaira nigricans*), blue shark (*Prionace glauca*), Atlantic bonito (*Sarda sarda*), smooth hammerhead (*Sphyrna zygaena*), southern bluefin tuna (*Thunnus maccoyii*), Atlantic bluefin tuna (*Thunnus thynnus*), and swordfish (*Xiphias gladius*).

Biomass data were sourced from Sea Around Us (SAU), which utilizes ICCAT data and the CMSY method - [@sau](https://www.seaaroundus.org/cmsy-method).
SAU estimates, available for 2010–2018, were averaged for this analysis. The combined biomass of the 18 key stocks in the Atlantic, which may enter the Saint Helena EEZ, totaled 5,503,844.45 tons.

The scaling factor  \(\lambda\) was calculated by dividing the global migratory biomass by SEAPODYM's skipjack and albacore stock estimates. This factor was then applied to SEAPODYM’s local skipjack and albacore biomass estimates to calculate total migratory biomass averaged by decade, model, and SSP. The annual proportion of Atlantic migratory biomass entering the Saint Helena domain was determined by dividing the scaled local biomass by the scaled Atlantic migratory biomass.

An increase in the proportion of migratory fish migrating to Saint Helena suggests that the island will become more suitable for tunas by the 2060s under these climate projections, except in the SSP1-GFDL model variant (`r table.ref("tab_tunas")`). This trend was integrated into StrathE2E to investigate its potential impacts on the entire food web and fisheries.

```{r bio-migratory, echo=FALSE, fig.align='center', fig.cap="Scaled annual mean migratory fish biomass from SEAPODYM model outputs for each decade (2010–2019 to 2060–2069), under two climate scenarios (SSP1-2.6 and SSP3-7.0) and two forcing models (CNRM and GFDL).", out.width = '100%'}
knitr::include_graphics("img/biomass_scaled_St_Helena_annual.png")
```

To assess the proportion of peak biomass that remained within the model domain even in the emigration period we extracted the maximum and minimum values from the seasonal cycle. Compared to the literature available, yellowfin tuna exhibit high fidelity to the area and remain in St. Helena waters for extended periods due to elevated productivity levels associated with the island and seamounts [@wright19; @wright21].

The full set of parameters used for migratory fish event timing is detailed in lines 17–26 of the event_timing_SAINT_HELENA_*.csv files for each decade under the four climate projections.

```{r tunas_cap}

tab_tunas_cap <- table_nums(name = "tab_tunas", 
    caption = "Time-variant biological event timing parameters related to migratory fish for the baseline period (2010-2019) and last decade of analysis (2060-2069) across the four climate projections, used as parameters data.")
```

`r table_nums("tab_tunas")`
